---
title: "data_wrangling_1_lec6"
output: github_document
date: "2024-09-29"
---

### load packages
```{r setup, warning=FALSE}
library(tidyverse)
```

### read data
```{r}
options(tibble.print_min = 3)
litters_df = read_csv("./data/FAS_litters.csv",
                      na=c("NA","."))
litters_df = janitor::clean_names(litters_df)
head(litters_df)

```

### select, select helper functions, rename, relocate
```{r}
#specify names of columns to keep 
select(litters_df, 
       group, 
       litter_number,
       gd0_weight,
       pups_born_alive)

#specify a range of columns to keep 
select(litters_df, 
       group:gd_of_birth)

#specify names of columns to remove
select(litters_df, 
       -pups_survive,
       -gd0_weight)

#rename 'selected' names of columns 
select(litters_df, 
       GROUP=group, 
       LiTter_NuMbEr=litter_number)

#rename only - not 'selecting' 
rename(litters_df, 
       GROUP=group, 
       LiTtER_NuMbEr=litter_number)

#select helper functions : starts_with()
select(litters_df, 
       starts_with("gd"))

select(litters_df, 
       ends_with("weight")
       )
select(litters_df, 
       contains("pups"))

#everything(): reorganize columns without discarding any columns
select(litters_df, 
       litter_number, 
       pups_survive, 
       everything())

#relocate - similar everything() select helper function 
relocate(litters_df, 
         litter_number,
         pups_survive)

#pull a single variable 
#'vector' format
pull(litters_df, 
     group)

#outputs one column - in 'dataframe' format
select(litters_df, 
       group)
```

la:
```{r}
options(tibble.print_min = 3)
pups_df = read_csv("./data/FAS_pups.csv",
                      na=c("NA","."))

pups_df = janitor::clean_names(pups_df)
colnames(pups_df)

select(pups_df, 
       litter_number, 
       sex, 
       pd_ears)
                      
```

## filter rows using logical expressions using filter()
comparison operators: >, >=, <, <=, ==, !=
presence or absence of values: %in%
find missing values: is.na()
logical operators: &, |, !

```{r drop_na}
#drop all rows with missing values
drop_na(litters_df)

#drop rows for which wt_increase is missing

drop_na(litters_df, 
        gd0_weight)
```

la:
```{r}
pups_df = read_csv("./data/FAS_pups.csv",
                      na=c("NA","."))

pups_df = janitor::clean_names(pups_df)
colnames(pups_df)

filter(pups_df, 
       sex==1)

filter(pups_df, 
       pd_walk<11, 
       sex==2)

```

## mutate
change variables
create new variales
```{r}
mutate(litters_df, 
       wt_gain = gd18_weight - gd0_weight, #create new variable 
       group = str_to_lower(group)) #update variable
```

la:
```{r}
mutate(pups_df, 
       pd_pivot_new = pd_pivot-7,
       pd_sum = pd_ears+pd_eyes+pd_pivot+pd_walk)
```


## arrange
rearrange columns 
```{r}
head(arrange(litters_df, 
             group, 
             pups_born_alive),
     5)
```

## multi-step data manipulation
piping: |> is "then" 
CMD+shift+M shortcut
output of one function call is used as first argument/input for next function call 

```{r}
litters_df = read_csv("./data/FAS_litters.csv", 
na=c("NA",".","")) |>
janitor::clean_names() |>
select(-pups_survive) |>
mutate(
wt_gain = gd18_weight - gd0_weight, 
group = str_to_lower(group))|>
drop_na(wt_gain)

litters_df 

```

placeholder "_": piping output goes to placeholder location
```{r}

litters_df |> 
  lm(wt_gain ~ pups_born_alive, data=_) |>
  broom::tidy()
```

limitations to pipe:
- too many sequences
- multiple inputs and outputs

la:
```{r}
pups_df = read_csv("./data/FAS_pups.csv",
                      na=c("NA",".")) |> 
  janitor::clean_names() |> 
  filter(sex==1) |> 
  select(-pd_ears) |> 
  mutate(pd_pivot_new = pd_pivot>7)

pups_df

```



