---
title: "data_wrangling_1_lec7"
output: github_document
date: "2024-09-29"
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
options(tibble.print_min=5)
```

## pivot_longer
```{r}
#read data
pulse_df = haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df

```

create multiple records for observation times
```{r}

pulse_tidy_df = pivot_longer(
  pulse_df, 
bdi_score_bl:bdi_score_12m, 
names_to="visit",
values_to="bdi")

pulse_tidy_df
```

remove bdi_score_prefix in 'visit' column
```{r}
pulse_tidy_df = pivot_longer(
  pulse_df,
  bdi_score_bl:bdi_score_12m, 
  names_to = "visit",
  names_prefix="bdi_score_",
  values_to="bdi")

pulse_tidy_df
```

don't create intermediate datasets
change bl to 00m
convert 'visit' variable to a factor variable

```{r}
pulse_df = haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |> 
  pivot_longer(
  bdi_score_bl:bdi_score_12m, 
  names_to = "visit",
  names_prefix="bdi_score_",
  values_to="bdi") |>
  mutate(
    visit = replace(visit, visit=="bl", "00m"),
    visit = factor(visit)
  )

pulse_df
```


## pivot_wider
tidy dataset
- good for further analysis
```{r}
analysis_results = 
  tibble(
    group = c("treatment",
              "treatment",
              "placebo",
              "placebo"
              ),
    time = c("pre",
             "post",
             "pre",
             "post"),
    mean = c(4, 8, 3.5, 4)
      
    )

analysis_results
```

untidy data using pivot_wider()
- good for human readability
- opposite of pivot_longer()
```{r}
pivot_wider(analysis_results, 
            names_from="time",
            values_from="mean")
```


## binding rows

create tidy dataset from untidy data spread across multiple tables
- create variable 'movie'
```{r}

#multiple tables
fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")

```

bind tables together
```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) |>
  janitor::clean_names() |>
  pivot_longer(
    female:male,
    names_to = "gender", 
    values_to = "words") |>
  mutate(race = str_to_lower(race)) |> 
  select(movie, everything()) 

lotr_tidy
```

## joining datasets

```{r}

pup_df = 
  read_csv(
    "./data/FAS_pups.csv",
    na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    sex = as.factor(sex)) 

litter_df = 
  read_csv(
    "./data/FAS_litters.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3) |>
  relocate(litter_number) |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))

fas_df = 
  left_join(pup_df, litter_df, by = "litter_number")

fas_df
```

